---
title: 'Class19: Genomic Analysis'
author: 'Yuhan Zhang (PID: A13829264)'
date: "12/2/2021"
output: pdf_document
---

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. **Hint**: The `read.table()`, `summary()` and `boxplot()` functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the medium value from saving the output of the `boxplot()` function to an R object and examining this object. There is also the `medium()` and `summary()` function that you can use to check your understanding.

```{r}
# read files
genome <- read.csv("rs8067378_ENSG00000172057.6.txt", sep = " ", 
                   row.names = 1, header = TRUE)
```


```{r}
library(dplyr)
genome %>% 
  group_by(geno)%>%
  summarize(med = median(exp))
```

The median expression level for A/A is 31.25, A/G is 25.06, G/G is 20.07.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
p <- ggplot(genome) + aes(geno, exp, fill = geno) + 
  geom_jitter(width = 0.25, alpha = 0.3) +
  geom_boxplot(notch = TRUE, alpha = 0.5) + 
  labs(x = "Genotype", y = "Expression")
p
```

The expression level with A/A SNP increase compare G/G. Thus SNP does effect the expression level of ORMDL3.