---
title: "Machine learning 1"
author: 'Yuhan Zhang (PID: A13829264)'
date: "10/22/2021"
output: pdf_document
---

# Clustering methods

Kmeans clustering in R is done with the `kmeans()` function.  
Here we make up some data to test and learning with.

```{r}
tmp <- c(rnorm(30, 3), rnorm(30, -3))
data <- cbind(x = tmp, y = rev(tmp))
plot(tmp)
```

Run `kmeans()` set k to 2 (i.e. the number of clusters you want) nstart 20(to tun multiple times). The thing with Kmeans is you have to tell it how many clusters you want.

```{r}
km <- kmeans(data, centers = 2, nstart = 20)
km
```

> Q. How many points are in each cluster?

```{r}
km$size
```

> Q. What ‘component’ of your result object details cluster assignment/membership?

```{r}
km$cluster
```

> Q. What ‘component’ of your result object details cluster center?

```{r}
km$centers
```

> Q. Plot x colored by the kmeans cluster assignment and
      add cluster centers as blue points
      
```{r}
plot(data, col = km$cluster)
points(km$centers, col = "blue", pch = 15, cex = 2)
```

# Hierarchical Clustering

We will use the `hclust()` funcitoin on the same data as before and see how this method works.

```{r}
hc <- hclust(dist(data))
hc
```

hclust has a plot method

```{r}
plot(hc)
abline(h = 7, col = 'red')
```

To find our membership vector we need to "cut" the tree and for this we use the `cutree()` unction and tell it the height to cut at.

```{r}
cutree(hc, h = 7)
```

We can also use `cutree()` and sate the number of k clusters we want ...

```{r}
grps <- cutree(hc, k = 2)
```

```{r}
plot(data , col = grps)
```

# Principal Component Analysis (PCA)

PCA is a super useful analysis method when you have lots of dimensions in your data...  

## PCA of UK food data

Import the data from a CSV file

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

How many rows and cols?
```{r}
dim(x)
```

```{r}
x[,-1]
```

```{r}
rownames(x) <- x[, 1]
x <- x[,-1]
x
```

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names = 1)
x
```

```{r}
barplot(as.matrix(x), col = rainbow(17))
```

```{r}
barplot(as.matrix(x), col = rainbow(17), beside = TRUE)
```

```{r}
mycols <- rainbow(nrow(x))
pairs(x, col = mycols, pch = 16)
```

## PCA to the rescue!

Here we will use the base R funcition for PCA, which is called `prcomp()`. This function requires us to transpose data.

```{r}
pca <- prcomp(t(x))
summary(pca)
```

```{r}
plot(pca)
```

We want score plot(a.k.a. PCA plot). Basically of PC1 vs PC2

```{r}
attributes(pca)
```

We are after the pca$x component for this plot...

```{r}
plot(pca$x[, 1:2])
text(pca$x[, 1:2], labels = colnames(x))
```

We can also examine the PCA "loadings", which tell us how much the original variables contribute to each new PC...

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot(pca$rotation[, 1], las = 2)
```

## One more PCA for today

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

> Q. How many genes and samples are in this data set?

```{r}
nrow(rna.data)
```

```{r}
colnames(rna.data)
```

```{r}
pca.rna <- prcomp(t(rna.data), scale = TRUE)
summary(pca.rna)
```

```{r}
plot(pca.rna)
```

```{r}
plot(pca.rna$x[, 1:2])
text(pca.rna$x[, 1:2], labels = colnames(rna.data))
```

Let's look at variance per PC
```{r}
pca.rna.var <- pca.rna$sdev^2

# Variance percentage
pca.rna.var.per <- round(pca.rna.var/sum(pca.rna.var)*100, 1)
pca.rna.var.per
```

Look at the variance using barplot:

```{r}
barplot(pca.rna.var.per, main="Scree Plot", 
        names.arg = paste0("PC", 1:10),
        xlab="Principal Component", ylab="Percent Variation")
```

We can use ggplot2 package to plot out our PCA:

```{r}
library(ggplot2)

df <- as.data.frame(pca.rna$x)

# Our first basic plot
ggplot(df) + 
  aes(PC1, PC2) + 
  geom_point()
```

We can label each sample and color each condition (either WT or KO)
```{r}
# Add a 'wt' and 'ko' "condition" column
df$samples <- colnames(rna.data) 
df$condition <- substr(colnames(rna.data),1,2)

p <- ggplot(df) + 
        aes(PC1, PC2, label=samples, col=condition) + 
        geom_label(show.legend = FALSE)
p
```

Add title and axis labels
```{r}
p + labs(title="PCA of RNASeq Data",
       subtitle = "PC1 clealy seperates wild-type from knock-out samples",
       x=paste0("PC1 (", pca.rna.var.per[1], "%)"),
       y=paste0("PC2 (", pca.rna.var.per[2], "%)")) +
     theme_bw()
```