---
title: "Class 10 Mini Project"
author: 'Yuhan Zhang (PID: A13829264)'
date: "10/29/2021"
output: 
  pdf_document:
    latex_engine: xelatex
always_allow_html: true
---

## 1. Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
rownames(candy) <- gsub("Õ", "'", rownames(candy))
head(candy)
```
### What is in the dataset?

> Q1. How many different candy types are in this dataset?

```{r}
candy.numType <- nrow(candy)
candy.numType
```


> Q2. How many fruity candy types are in the dataset?

```{r}
fruityNum <- sum(candy$fruity)
fruityNum
```

## 2. What is your favorate candy?

One of the most interesting variables in the dataset is `winpercent`. For a given candy this value is the percentage of people who prefer this candy over another randomly chosen candy from the dataset (what 538 term a matchup). Higher values indicate a more popular candy.

We can find the `winpercent` value for Twix by using its name to access the corresponding row of the dataset. This is because the dataset has each candy name as `rownames` (recall that we set this when we imported the original CSV file). For example the code for Twix is:

```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it’s `winpercent` value?

```{r}
candy["WelchÕs Fruit Snacks", ]$winpercent
```


> Q4. What is the `winpercent` value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```


> Q5. What is the `winpercent` value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

> Side-note: the skimr::skim() function  
There is a useful skim() function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
# install.packages("skimr")
library("skimr")
skim(candy)
```

Use `skim()` and answer the following:

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent is in a different scale to majority since it has standard deviation of 14.7.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

0: no chocolate
1: contain chocolate

A good place to start any exploratory analysis is with a histogram. You can do this most easily with the base R function `hist()`. Alternatively, you can use `ggplot()` with `geom_hist()`. Either works well in this case and (as always) its your choice.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
win.hist <- ggplot(candy, aes(x = winpercent)) + 
  geom_histogram()
win.hist
```


> Q9. Is the distribution of winpercent values symmetrical?

Yes the distribution is symmetrical

> Q10. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
```

The center is below 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
candy.chocolateType <- candy$winpercent[as.logical(candy$chocolate)]
candy.fruityType <- candy$winpercent[as.logical(candy$fruity)]

t.test(candy.chocolateType, candy.fruityType)
```

On average chocolate is higher ranked than fruity type.

> Q12. Is this difference statistically significant?

There is significant difference (p = 2.871e-08).  

## 3. Overall Candy Rankings

Let’s use the base R `order()` function together with `head()` to sort the whole dataset by `winpercent`. Or if you have been getting into the tidyverse and the **dplyr** package you can use the `arrange()` function together with `head()` to do the same thing and answer the following questions:

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy.leastFav <- candy %>% arrange(winpercent) %>% head(5)
candy.leastFav
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy.mostFav <- candy %>% arrange(desc(winpercent)) %>% head(5)
candy.mostFav
```

To examine more of the dataset in this vain we can make a barplot to visualize the overall rankings. We will use an iterative approach to building a useful visulization by getting a rough starting plot and then refining and adding useful details in a stepwise process.

```{r, fig.height = 10}
ggplot(candy, aes(winpercent, rownames(candy))) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height = 10}
ggplot(candy, aes(winpercent, reorder(rownames(candy), winpercent))) +
  geom_col()
```

### Time to add some useful color

Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

Now let’s try our barplot with these colors. Note that we use `fill=my_cols` for `geom_col()`. Experement to see what happens if you use `col=mycols`.

```{r, fig.height=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

Now, for the first time, using this plot we can answer questions like:

> Q17. What is the worst ranked chocolate candy?

sixlets

> Q18. What is the best ranked fruity candy?

starbusts

## 4. Taking a look at pricepercent

What about value for money? What is the the best candy for the least money? One way to get at this would be to make a plot of `winpercent` vs the `pricepercent` variable. The `pricepercent` variable records the percentile rank of the candy’s price against all the other candies in the dataset. Lower vales are less expensive and high values more expensive.

To this plot we will add text labels so we can more easily identify a given candy. There is a regular `geom_label()` that comes with ggplot2. However, as there are quite a few candys in our dataset lots of these labels will be overlapping and hard to read. To help with this we can use the `geom_text_repel()` function from the **ggrepel** package.

```{r}
# install.packages("ggrepel")
library(ggrepel)

# How about a plot of price vs win
my_cols[as.logical(candy$fruity)] <- "red"
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Chocolate (Reese's miatures at the right bottom corner)

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Nik L Nip

> **opitional** Q21. Make a barplot again with `geom_col()` this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping `geom_col()` for `geom_point()` + `geom_segment()`.

```{r, fig.height = 10}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## 5. Exploring the correlation structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the **corrplot** package to plot a correlation matrix.

```{r}
# install.packages("corrplot")
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate and fruity

> Q23. Similarly, what two variables are most positively correlated?

bar and chocolate

## 6. Principal Component Analysis

Let’s apply PCA using the `prcomp()` function to our candy dataset remembering to set the `scale=TRUE` argument.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

Now we can plot our main PCA score plot of PC1 vs PC2.

```{r}
plot(pca$x[,1:2])
```

We can change the plotting character and add some color:

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

Using ggplot2:

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

Again we can use the **ggrepel** package and the function `ggrepel::geom_text_repel()` to label up the plot with non overlapping candy names like. We will also add a title and subtitle like so:

```{r, fig.width = 8}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

more candy labels you can change the `max.overlaps` value to allow more overlapping labels or pass the ggplot object `p` to **plotly** like so to generate an interactive plot that you can mouse over to see labels:

```{r}
library(plotly)
ggplotly(p)
```

Let’s finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of `chocolate` and `fruity` and the similar effects of `chocolate` and `bar` (i.e. we already know they are correlated).

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, hard, pluribus, yes because fruity candy are usually hard candy or come with pluribus, and fruity candy usually are not chocolate / caramel / nuty / crisp