---
title: 'Class17: COVID-19 Vaccination Rates'
author: 'Yuhan Zhang (PID: A13829264)'
date: "11/24/2021"
output: github_document
---

# Background



Import the data:

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv", header = TRUE)
head(vax)
```

> Q. How many entries do we have?

```{r}
nrow(vax)
```

We can use the **skimr** package and the `skim()` function to get a quick overview of structure of this dataset.

```{r}
skimr::skim(vax)
```

Notice that one of these columns is a date column. Working with time and dates get's annoying quickly. we can use the **lubridate** package to make this easy...

```{r}
library(lubridate)

today()
```

> Q. How many days since the first entry in the dataset?

```{r}
vax$as_of_date[1]
```

This will not work because our data column was read as character
```{r}
# today() - vax$as_of_date[1]
```

```{r}
d <- ymd(vax$as_of_date)
```

```{r}
today() - d[1]
```

I will make the `as_of_date` column Date format...

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

> Q. When was the data last updated? What it is the last date in this dataset? How many days since the last update?

```{r}
today() - vax$as_of_date[nrow(vax)]
```

> Q. How many days does the dataset span?

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q. How many different ZIP code areas are in this dataset?

```{r}
length(unique(vax$zip_code_tabulation_area))
```

To work with ZIP codes we can use the **zipcodeR**

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

Calculate distance:
```{r}
zip_distance('92037','92109')
```

Pull census data from zipcode:
```{r}
reverse_zipcode(c('92037', "92109") )
```

## Focus on San Diego County

We want to subset the full CA `vax` data down to just San Diego County

We could do this with base R
```{r}
inds <- vax$county == "San Diego"
nrow(vax[inds,])
```

Subsetting can get tedious and complicated quickly when you have multiple things we want to subset by.

```{r message=FALSE}
library(dplyr)
```

We will use the `filter()` function to do our subsetting from now on.

We want to focus on San Diego County

```{r}
sd <- filter(vax, county == "San Diego")
nrow(sd)
```

More complicated subsetting...

```{r}
sd.20 <- filter(vax, county == "San Diego", 
       age5_plus_population > 20000)
nrow(sd.20)
```

> Q. What is the average vaccination rate of San Diego count as of yesterday?

```{r}
sd.now <- filter(vax, county == "San Diego", as_of_date == "2021-11-23")
head(sd.now)
```

```{r}
summary(sd.now$percent_of_population_fully_vaccinated)
```

> Q. Make a histogram of these values.

```{r}
hist(sd.now$percent_of_population_fully_vaccinated, 
     xlab = "% Population", main = "Percent of Population Fully Vaccinated")
```

This is going to be susceptible to being skewed by ZIP code areas with small populations. These will have big effects for just a small number of unvax-ed folks...

> Q. What is the population of the 92037 ZIP code area?

```{r}
lj <- filter(sd.now, zip_code_tabulation_area == 92037)
lj$age5_plus_population
```

> Q. What is the average vaccination value for this UCSD/La Jolla ZIP code area?

```{r}
lj$percent_of_population_fully_vaccinated
```

```{r}
lj2 <- filter(sd.now, zip_code_tabulation_area == 92122)
lj2$age5_plus_population
```

```{r}
lj2$percent_of_population_fully_vaccinated
```

```{r}
filter(sd.now, zip_code_tabulation_area == 92124)
```

--- 

Time series of vaccination rate for a given ZIP code area.
Start with 92037.

```{r}
ucsd <- filter(vax, zip_code_tabulation_area == 92037)
head(ucsd)
```

```{r}
library(ggplot2)

ggplot(ucsd) + 
  aes(as_of_date, percent_of_population_fully_vaccinated) + 
  geom_line(group = 1) + 
  geom_point() + 
  labs(title = paste0("Vaccination Rate of La Jolla, 92037 since ", 
                      ucsd$as_of_date[1]),
       x = "Date", y = "Percent Rate")
```

Let's make this plot for all San Diego Count ZIP code areas that have a population as least as large as 92037.

```{r}
sd.36 <- filter(vax, county == "San Diego", 
             age5_plus_population > 36144)
head(sd.36)
```

> How many ZIP code areas in San Diego County have a population larger than 92037?

```{r}
length(unique(sd.36$zip_code_tabulation_area))
```

```{r}
ggplot(sd.36) + 
  aes(as_of_date, percent_of_population_fully_vaccinated, 
      group = zip_code_tabulation_area) + 
  geom_line() + 
  labs(title = 
         paste0(
           "Vaccination Rate of San Diego County population \n greater than zipcode 92037 since ", 
           ucsd$as_of_date[1]),
       x = "Date", y = "Percent Rate")
```

> Q. Make a plot like this for the all ZIP code areas in the State with a population at least as large as La Jolla.

```{r}
ca <- filter(vax, age5_plus_population > 36144)
```

```{r}
length(unique(ca$zip_code_tabulation_area))
```

```{r}
ggplot(ca) + 
  aes(as_of_date, percent_of_population_fully_vaccinated, 
      group = zip_code_tabulation_area) + 
  geom_line(alpha = 0.3) + 
  labs(title = 
         paste0(
           "Vaccination Rate of San Diego County population \n greater than zipcode 92037 since ", 
           ucsd$as_of_date[1]),
       x = "Date", y = "Percent Rate") + 
  geom_hline(yintercept = 0.67, col = "red")
```

> Q. What is the mean across the state for these 36k + population areas?

```{r}
ca.now <- filter(ca, as_of_date == "2021-11-23")
summary(ca.now$percent_of_population_fully_vaccinated)
```