---
title: 'Class 06: R Functions'
author: 'Yuhan Zhang (PID: A13829264)'
date: "10/15/2021"
output: pdf_document
---

## Quick Rmarkdown intro

We can write text of course just like any file. We can **style text to be bold** or *italic*.

Do:

- this
- and that
- and another thing

This is more text  
and this is a new line

---

We can include some code:

```{r}
plot(1:10)
```


```{r}
# This is a comment and will not be passed to R
plot(cars)
```


## Time to write the funciton

> **Q1**. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]


```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90) 
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80) 
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

First I want to find the lowest score. I can use the **min()** to find it and the **which.min()** funciton to find where it si (i.e. its position in the vector)

```{r}
which.min(student1)
```
I can use minus to get everything in the vector but hte lowest score.
```{r}
student1[-which.min(student1)]
```

Now i can call the **mean()** function to get the average.

```{r}
mean(student1[-which.min(student1)])
```

Does this work for student2?
```{r}
mean(student2[-which.min(student2)])
```

NO!
Why not?

```{r}
student2
which.min(student2)
```

```{r}
mean(student2, na.rm = TRUE)
```

```{r}
student3
```

One great idea is to replace the NA values with zero. Let's do it.

The internet gods told me to try this

```{r}
which(is.na(student2))
```

The is.na() function returns a logical vector where TRUE elements indicate the presence of NA values.
```{r}
is.na(student2)
```

```{r}
student2[is.na(student2)]
```

Let's replace NAs with zero
```{r}
student.prime <- student2
student.prime[is.na(student2)] = 0
student.prime
```

```{r}
x <- 1:5
x
x[2] = 100
x
x[3] = 200
x
```


Ok we are so so close let's put these bits together to get our mean excluding the lowest score.
```{r}
student.prime <- student2
student.prime[is.na(student2)] = 0
mean(student.prime[-which.min(student.prime)])
```

How about student3?

```{r}
student.prime <- student3
student.prime[is.na(student3)] = 0
mean(student.prime[-which.min(student.prime)])
```

Great! We got it. this works. Let's simplify and make as clear as we can.  
  
We can make the object names more clear

```{r}
x <- student3
x[is.na(x)] = 0
mean(x[-which.min(x)])
```

Doh! eeijt Barry entered the data wrong...

```{r}
student4 <- c(100, NA, 90, "90", 90, 90, 97, 80) 
```

```{r}
x <- student4
x <- as.numeric(x)
x[is.na(x)] = 0
mean(x[-which.min(x)])
```

Now finally we can write our function:
All functions have at least 3 things:
A name, input args, and a body.

```{r}
grade <- function(x) {
  x <- as.numeric(x)
  x[is.na(x)] = 0
  mean(x[-which.min(x)])
}
```


```{r}
grade(student1)
grade(student2)
grade(student3)
```

### Now grade a whole class

First we got to read the gradebook for the class

```{r}
url <- "https://tinyurl.com/gradeinput"
scores <- read.csv(url, row.names = 1)
scores
```
We are going to use the supper useful **apply()** funciton to grade all the students with our **grade()** function.

```{r}
ans <- apply(scores, 1, grade)
```

> **Q2**. Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]

```{r}
ans[which.max(ans)]
```

> **Q3**. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

Here I will use the **apply()** function again but this time look at the columns, which represent different homeworks.

```{r}
apply(scores, 2, mean)
```

I can ignore the NA missing values with na.rm = TRUE

```{r}
apply(scores, 2, mean, na.rm = TRUE)
```

Replace or mask NA values to zero
```{r}
mask <- scores
is.na(mask)
```

```{r}
mask <- scores
mask[is.na(mask)] = 0
mask
```

Now we can use apply on our "masked" scores
```{r}
apply(mask, 2, mean, na.rm = TRUE)
```

> **Q4**. Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

Here we will use the **cor()**
```{r}
cor(mask$hw5, ans)
```

I can call the **cor()** for every homework and get a value for each but that sucks. Let's use **apply()** and do them all in one go.

```{r}
apply(mask, 2, cor, y = ans)
```

Make a boxplot
```{r}
boxplot(scores)
```

Edit to optionally drop the lowerest score:
```{r}
grade2 <- function(x, drop.lowest = TRUE) {
  x <- as.numeric(x)
  x[is.na(x)] = 0
  if(drop.lowest) {
    mean(x[-which.min(x)])
  } else {
    mean(x)
  }
}
```

```{r}
student <- c(100, 100, 100, 90)
grade2(student, drop.lowest = FALSE)
```




